---
title: "ENVS_193DS_Final"
author: "Jason Han"
date: 6/11/2025
format: html
editor: visual
execute:
  warning: false
  error: false
  message: false
toc: true
toc-depth: 5
---

##### githubrepo:

##### https://github.com/ButterEggs/ENVS-193DS_spring-2025_final

### Set up chunk for Problem 2 and 3

```{r}
suppressPackageStartupMessages({
  library(tidyverse) #reading in packages
  library(dplyr)
  library(here)
  library(gt)
  library(janitor)
  library(ggplot2)
  library(DHARMa)
  library(MuMIn)
  library(ggeffects)}) 

sst <- read_csv(here("data", "SST_update2023.csv"), show_col_types = FALSE) #reading in sst data

nest_boxes <- read_csv(here("data", "occdist.csv")) #reading in nest_boxes data
```

### Problem 1. Research writing

### a. Transparent statistical methods

In part 1, they used a Pearson's correlation test. In part 2, they used a ANOVA test.

### b. More information needed

In part 2, the one-way ANOVA test compares the nitrogen loads from five different sources showing how at least one group differs from the others, but not which ones. This is shown with a statistically significant result of (p = 0.02). With this information, we need to conduct a Tukey’s Honestly Significant Difference (HSD) test in order to identify which specific sources, such as fertilizer or urban runoff, have significantly different mean nitrogen loads. This would help clarify the biological relevance of the results by showing which sources are driving the overall difference. Furthermore, the p-value tells us that the result is statistically significant, but doesn’t indicate how much variance in the nitrogen load is explained by the type of source. This is why an effect size for ANOVA would help add context to assess if the differences between the sources are biologically meaningful.

### c. Suggestions for rewriting

i.  We found a (insert effect size) relationship between distance from headwater (km) and annual total nitrogen load (kg year⁻¹), (Pearson’s correlation test: r = correlation coefficient, p = 0.03, α = significance level)

ii. We found that there is a (insert effect size) difference in average nitrogen load (kg year⁻¹) between source: urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands (one-way ANOVA, F(4, df_within) = F-statistic, p = 0.02, α = significance level).

### Problem 2. Data visualization

### a. Cleaning and summarizing

```{r}
sst_clean <- sst |> #cleaning sst data
  clean_names() |> 
  mutate(
    date = ymd(date), #convert 'date' column to Date format
    year = factor(year(date)),  #took year from 'date' converting into factor
    month = factor(month(date, label = TRUE, abbr = TRUE), #taking the month and converting into factor
                   levels = month.abb, 
                   ordered = TRUE)) |>   # ordered Jan–Dec
  filter(year %in% c("2018", "2019", "2020", "2021", "2022", "2023")) |> #keep only data from 2018 to 2023 
  droplevels() |>  #dropped unused factor levels for 'year' and 'month'
  group_by(year, month) |>  #grouped data by year and month
  summarize(mean_monthly_sst = mean(temp, na.rm = TRUE), .groups = "drop") #calculate the mean sea surface temperature for each group
  
slice_sample(sst_clean, n = 5) #previewing a random sample of 5 rows from the cleaned dataset
str(sst_clean) #displaying the structure of the cleaned and summarized dataset
```

### b. Visualize the data

```{r message=FALSE, warning=FALSE}
ggplot(sst_clean, aes(x = month, y = mean_monthly_sst, group = year, color = year)) + #creating a line plot of mean monthly sea surface temperature by month, grouped and colored by year

  geom_line(linewidth  = 1.1) +  #added lines for each year with thicker line width
  geom_point(size = 2) + #added points at each month to show individual data values
  
  labs(
    x = "Month", #labeled x-axis 
    y = "Mean monthly sea surface temperature (°C)", #labeled y-axis 
    color = "Year") +

  scale_color_manual( 
    values = colorRampPalette(c("#a6cee3", "#08306b"))(6)) +  #used a custom color gradient (from light blue to dark blue)

  theme_classic(base_size = 13) +
  theme(
    axis.ticks.y = element_line(color = "black"), #added black ticks on y-axis
    axis.line.y = element_line(color = "black"), #added black axis line on y-axis   
    legend.position = c(0.15, 1.0), #moved legend to top-left inside the plot
    legend.justification = c(0, 1), #anchored legend to top-left corner
    legend.background = element_blank(), #removed background fill from legend
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank(), #removed grid lines
    #setting size for axis text, axis title, legend title, and legend text.
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 13),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

### Problem 3. Data analysis

```{r echo=FALSE}
nest_boxes_clean <- nest_boxes |> #cleaning nest_boxes data
  clean_names()
```

### a. Response variable

In this dataset, the columns sp, cs, e, and tm are binary response variables representing if nest boxes are occupied by specific species. A value of 1 means the nest box was occupied by that species for example a swift parrot, while a 0 means it was either occupied by a different species or left empty.

### b. Purpose of study

The Swift Parrots are the target species for nest box restoration, while the others are competitors that may exploit the boxes. The Common Starlings are invasive and although the Tree Martins are native they both pose a threat towards the restoration effort by outcompeting the endangered parrots for nesting sites.

### c. Difference in “seasons”

The two seasons in the study are 2016 and 2019. In 2016, the nest boxes had just been deployed, whereas by 2019, they had been established for three years, allowing researchers to assess how box age influenced occupancy patterns and competitor behavior over time.

### d. Table of models

| model number | season | distance to forest edge | model description |
|----|----|----|----|
| 0 |  |  | No predictors (null model) |
| 1 | X | X | Year + Distance (saturated model) |
| 2 | X |  | Year |
| 3 |  | X | Distance to forest edge |
|  |  |  |  |

### e. Run the models

```{r}
model0 <- glm( # Null model
  sp ~ 1, 
  data = nest_boxes_clean,
  family = binomial())

model1 <- glm( # Full model (saturated)
  sp ~ season + edge_distance, 
  data = nest_boxes_clean, 
  family = binomial())

model2 <- glm( # Model with only season as a predictor
  sp ~ season,
  data = nest_boxes_clean,
  family = binomial())

model3 <- glm( # Model with only distance to forest edge as a predictor
  sp ~ edge_distance,
  data = nest_boxes_clean,
  family = binomial())
```

### f. Check the diagnostics

```{r}
#displaying the diagnostic plots for each model
Model0 <- plot(simulateResiduals(model0)) #plot for null model

Model1 <- plot(simulateResiduals(model1)) #plot for saturated model

Model2 <- plot(simulateResiduals(model2)) #plot for season predictor 

Model3 <- plot(simulateResiduals(model3)) #plot for distance from forest edge predictor
```

### g. Select the best model

```{r}
AICc(model1, #compute AICc (Akaike Information Criterion) for each model
     model2,
     model3,
     model0) |> 
  arrange(AICc) #arrange the models in ascending order of AICc
```

The best model, as determined by Akaike’s Information Criterion (AIC), includes season and distance from the forest edge as predictors of Swift Parrot nest box occupancy and has the lowest AIC value of 226.31.

### h. Visualize the model predictions

```{r}
model_prediction <- ggpredict(model1, terms = c("edge_distance [all]", "season")) |> #generated model predictions from the best model (model1) across all values of edge_distance, grouped by season
  rename(season = group) #converting season to a factor 

nest_boxes_clean <- nest_boxes_clean |> 
  mutate(
    sp = as.numeric(sp),  #making sure the response variable is numeric
    season = factor(season))

#creating the ggplot visualization
ggplot(data = nest_boxes_clean, 
       aes(x = edge_distance, y = sp, color = season)) +
  
  geom_point(alpha = 0.45, size = 2) +  #adding raw data points with transparency

  geom_ribbon(data = model_prediction, #adding 95% confidence intervals from model predictions
              aes(x = x, ymin = conf.low, ymax = conf.high, fill = season),
              alpha = 0.2, inherit.aes = FALSE) +

  geom_line(data = model_prediction, #adding model predicted probability lines
            aes(x = x, y = predicted, color = season),
            linewidth = 1.1, inherit.aes = FALSE) +

  scale_y_continuous(limits = c(0, 1), breaks = c(0, 0.5, 1)) + #y-axis scaled for probability
  
  #adding custom colors and fill  for each season
  scale_fill_manual(values = c("2016" = "#a6cee3", "2019" = "#08306b")) +
  scale_color_manual(values = c("2016" = "#a6cee3", "2019" = "#08306b")) +

  #adding a minimal theme and adjusting appearance
  theme_minimal(base_size = 14) +
  theme(
    panel.background = element_blank(),
    panel.grid = element_blank(),
    axis.ticks = element_line(color = "black"),
    axis.line = element_line(color = "black"),
    legend.position = c(0.9, 0.6)) +

  labs( #adding labels 
  title = "Swift Parrot Nest Box Use by Distance from Forest Edge",
  x = "Distance from forest edge (m)",
  y = "Probability of nest box occupancy (%)",
  fill = "Season", 
  color = "Season")

```

### i. Write a caption for your figure.

***Figure 1: Probability of nest box occupancy by Swift Parrot decreases as distance from forest edge (m) increases.*** Data collected by Australian National University, available in the “occdist.csv” dataset (Stojanovic, Dejan; Owens, Giselle; et al., 2021), DOI: https://doi.org/10.5061/dryad.83bk3j9sb. The points show the raw data for occupancy of Swift Parrots in nest boxes (n = 49) and also nest boxes unoccupied by Swift Parrots (n = 178). Predicted probabilities of Swift Parrots occupying nest boxes across varying distances from the forest edge in meters. The shaded ribbons indicate 95% confidence intervals surrounding the predicted values generated by the model including both season and edge distance. The 2016 season is shown in light blue, while blue corresponds to 2019.

### j. Calculate model predictions

```{r}
ggpredict(model1, #calculating the predicted probaility at 0m and 900m for 2016
          terms = c("edge_distance [0, 900]", "season [2016]"))

ggpredict(model1, #calculating the predicted probaility at 0m and 900m for 2016
          terms = c("edge_distance [0, 900]", "season [2019]"))
```

### k. Interpret your results

Swift Parrots are more likely to use nest boxes that are positioned closer to the forest edge, as shown in the figure from part h. The model predictions from part j show that in 2016, the probability of occupancy at 0 meters from the forest edge is 0.48 (95% CI: \[0.33, 0.64\]), while at 900 meters it drops to 0.12 (95% CI: \[0.06, 0.24\]). For the 2019 season, the likelihood of occupancy at 0 meters is 0.30 (95% CI: \[0.18, 0.44\]), and only 0.06 (95% CI: \[0.03, 0.13\]) at 900 meters. These results indicate that Swift Parrots prefer nest boxes near forest edges, likely due to better access to foraging habitat and natural cavities, while also avoiding areas with higher densities of competitors. The results also suggest that Swift Parrots may be avoiding competition from Tree Martins, which show the opposite pattern, as their probability of nest box occupancy increases with greater distance from the forest edge.

### Problem 4a. Comparing visualizations

i.  Looking at my affective visualization on homework 3 compared to my exploratory visualization on homework 2, my affective visualization conveys a more representational meaning. The drawings, pictures, and symbols convey the variables that influenced the visuals of my data but is not accurate or to-scale. On the other hand my exploratory visualization uses accurate to-scale plots that represent my data and is meant for analysis.

ii. Looking at both types of visualizations, they both convey the same data which is steps vs school day and also temperature. Both visualizations emphasize how the steps variable is influenced by either the type of day or temperature.

iii. With only 3 observations in my exploratory visualization since I just started to collect data, I only had the mean steps for school days which was about 7000 steps. Similarly for the steps vs temperature plot, no trends could be made since there was only 3 observations. For the affective visualization, I made the assumption that steps were higher on non-school days than school days, in addition to having all the other variables pointing towards that assumption.

iv. Feedback on my visualization included adding a some type of representation for what a school day looked like compared to a non-school day. With this feedback, I included a photo of Storke tower to represent a school day and a photo of my workplace to represent my non-school days. I also received feedback about incorporating my actual data making my visualization to-scale, although I didn't do this since it was just supposed to be more of a representation.